Multivariate Density Estimation: Comparing Transformation Random Forest, Normalizing Flows and Copulas

### 2025-09-30 report/chapter03.Rnw
.

\begin{table}[htbp]
  \centering
  \textit{(average LL; nats per example).}
  \caption{UCI: average test log-likelihood per example (nats; higher is better). Baselines (first seven rows): means $\pm$ 2SE as reported by \citet{papamakarios2017masked}. TRTF (ours): single-seed measurements at $N=2500$ (no SE). Entries marked ``--'' indicate configurations not executed in this draft.}
  \label{tab:uci-loglik}
  \begin{tabular}{lrrrr}
    \hline
    Model & POWER & GAS & HEPMASS & MiniBooNE \\
    \hline
    Gaussian          & $-7.74 \pm 0.02$ & $-3.58 \pm 0.75$ & $-27.93 \pm 0.02$ & $-37.24 \pm 1.07$ \\
    MADE              & $-3.08 \pm 0.03$ & $ 3.56 \pm 0.04$ & $-20.98 \pm 0.02$ & $-15.59 \pm 0.50$ \\
    MADE MoG          & $ 0.40 \pm 0.01$ & $ 8.47 \pm 0.02$ & $-15.15 \pm 0.02$ & $-12.27 \pm 0.47$ \\
    Real NVP (5)      & $-0.02 \pm 0.01$ & $ 4.78 \pm 1.80$ & $-19.62 \pm 0.02$ & $-13.55 \pm 0.49$ \\
    Real NVP (10)     & $ 0.17 \pm 0.01$ & $ 8.33 \pm 0.14$ & $-18.71 \pm 0.02$ & $-13.84 \pm 0.52$ \\
    MAF (5)           & $ 0.14 \pm 0.01$ & $ 9.07 \pm 0.02$ & $-17.70 \pm 0.02$ & $-11.75 \pm 0.44$ \\
    MAF MoG (5)       & $ 0.30 \pm 0.01$ & $ 9.59 \pm 0.02$ & $-17.39 \pm 0.02$ & $-11.68 \pm 0.44$ \\
    TRTF (ours)    & $-7.17 \pm 0.39$ & $-2.41 \pm 0.37$ & $-25.47 \pm 0.37$ & $-30.01 \pm 1.26$ \\
    \hline
  \end{tabular}
\end{table}

\paragraph{MiniBooNE.} Table~\ref{tab:uci-loglik} shows that the Gaussian reference yields $-37.24 \pm 1.07$ nats, providing a weak baseline. MADE reaches $-15.59 \pm 0.50$ nats, the Real NVP variants lie near $-13.7$ nats, and MAF MoG improves to $-11.68 \pm 0.44$ nats. Our TRTF result attains $-30.01 \pm 1.26$ nats at $N=2500$, improving over the Gaussian baseline yet trailing the flow families by a wide margin. This ranking is consistent with the separable Jacobian and the forest aggregation discussed in Section~\ref{sec:models-implementation}. The high dimensionality of MiniBooNE amplifies residual misfit through the triangular determinant.\;Clipping: validation-tuned bound $H$ applied; the exact value is recorded with the experiment logs.

\paragraph{POWER.} POWER offers a milder conditional structure and lower dimensionality. Table~\ref{tab:uci-loglik} reports that TRTF records $-7.17 \pm 0.39$ nats at $N=2500$, which falls short of the flow baselines. Real NVP with ten steps reaches $0.17 \pm 0.01$ nats, while MAF MoG attains $0.30 \pm 0.01$ nats. The gap indicates that the current TRTF configuration underutilizes structure in this benchmark; additional seeds or hyperparameter tuning may recover the performance previously observed at smaller sample sizes.\;Clipping: validation-tuned bound $H$ applied; the exact value is recorded with the experiment logs.

\paragraph{GAS and HEPMASS.} The TRTF results on GAS and HEPMASS yield $-2.41 \pm 0.37$ and $-25.47 \pm 0.37$ nats, respectively. Both scores remain below the flow baselines, emphasizing that the present configuration sacrifices likelihood accuracy for interpretability. Additional seeds and tuning remain planned, yet we retain the current numbers to document the outcome of the standardized pipeline at $N=2500$.\;Clipping: validation-tuned bound $H$ applied; the exact values are recorded with the experiment logs.

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.85\linewidth]{figure/N_sensitivity_all.png}
  \caption{Test negative log-likelihood (NLL; nats; lower is better) versus sample size $N$ on the UCI benchmarks. Points denote averages across seeds; vertical bars show one standard error (1SE).}
  \label{fig:n-sensitivity}
\end{figure}

@@

\paragraph{Interpretation.} The real-data evidence aligns with the synthetic diagnostics in Section~\ref{sec:synthetic-results}. MiniBooNE exposes the limits of separable structure in high dimensions, and the updated POWER value shows that the present TRTF configuration no longer matches flow baselines once the training budget increases to $N=2500$. GAS and HEPMASS also trail the published flows, illustrating that interpretability and exact inversion come at a likelihood cost under the current hyperparameters. Table~\ref{tab:real-compute} documents the corresponding compute budgets and confirms the anticipated near-linear growth in wall-clock time.

\section{Reproducibility}\label{sec:reproducibility}
We avoid AIC or BIC because effective parameter counts differ across estimators, and we do not treat small likelihood differences as practically significant when $\pm 2$ SE intervals overlap. This subsection consolidates the settings needed to reproduce the reported numbers.

- Data splits and direction
  - Synthetic: fixed train/validation/test proportions $0.60/0.20/0.20$; evaluations use the shared direction $S:u\to z$ in standardized coordinates and apply the diagonal affine correction once for reporting.
  - Real data: use official splits where provided (MiniBooNE) and the same standardized evaluation pipeline; otherwise adopt the same $0.60/0.20/0.20$ convention.

- Seeds
  - Synthetic generators and model fits: seeds $\{11,13,17,19,23\}$ across repeats; permutation studies average over all $4!=24$ orderings in the 4D case.
  - Real data (UCI + MiniBooNE): single-seed runs with seed $42$ for training/evaluation in this draft.

- Standardization and evaluation
  - Standardize features with training-split $(\mu,\sigma)$ only; compute all derivatives/Jacobians in $u$; report on $x$ via the affine correction in Eq.~\eqref{eq:transport-affine}.
  - TRTF uses forest aggregation and monotone CDF smoothing so that the induced likelihood matches the separable triangular form (Sec.~\ref{sec:transport-trtf}).

- Hyperparameters and tuning
  - TTM-Sep: monotone one-dimensional bases for $h_k$ (identity, integrated sigmoids, softplus-like edge terms, integrated RBFs); low-degree polynomial features for $g_k$; ridge regularization on all coefficients; log-derivative clipping to $[-H,H]$ (bound $H$ tuned on validation). Degree and penalty strengths are selected by validation; ordering is fixed to the natural order in headline tables and varied in robustness checks.
  - TTM-Sep: monotone one-dimensional bases for $h_k$ (identity, integrated sigmoids, softplus-like edge terms, integrated RBFs); low-degree polynomial features for $g_k$; ridge regularization on all coefficients; log-derivative clipping to $[-H,H]$ (bound $H$ tuned on validation). Degree and penalty strengths are selected by validation; ordering is fixed to the natural order in headline tables and, when heuristics are enabled, chosen as the better of identity vs. Cholesky-pivoted (with optional Gaussianization) according to validation NLL.
  - TRTF: forest aggregation with strictly increasing conditional CDFs after standard monotone smoothing; remaining fit options follow package defaults unless stated; we record the number of trees, depth and split rules in the experiment logs.
  - Copulas (diagnostics only for $K\le 3$): probit pseudo-observations and kernel density copula via \texttt{kdecopula} with default bandwidth selection; independence and Gaussian baselines are used only for reference in text where noted.
  - Exact choices (e.g., basis sizes, ridge penalties, selected $H$) are captured alongside each run in the experiment logs and summarized inline where relevant; we avoid duplicate tables in the PDF.

Final safeguard settings used for the reported results. For Half-Moon ($n=250$) and 4D ($n=250$), TTM-Sep used degree$_g=2$, ridge $\lambda=0$, and no log-derivative clipping was activated (no terms hit the bound). The $n=25$ 4D case overflowed under $\lambda=0$; reruns with $\lambda>0$ and tighter $H$ removed the failure but are omitted as out of scope. Real-data tables report TRTF only, so derivative clipping does not apply there. Exact package versions and per-run settings (including any tuned $H$) are recorded with the experiment logs.

- Software and hardware
  - R with packages: \texttt{tram}, \texttt{trtf}, \texttt{partykit}, \texttt{mlt}, \texttt{dplyr}, \texttt{parallel}, and \texttt{knitr}/LaTeX for the report. We record package versions via \texttt{sessionInfo()} in run logs.
  - Single-threaded BLAS by default; optional parallel training for TRTF via \texttt{options(trtf.train\_cores = 4)} when available.
  - CPU-only runs on a laptop-class machine; logs include hardware notes (CPU model, RAM) and wall-clock timings (Table~\ref{tab:real-compute}).

>>>>

\begin{table}[htbp]
 \centering
 \caption{UCI: average test log-likelihood per example (higher is better). Baselines (first seven rows): means $\pm$ 2SE as reported by \citet{papamakarios2017masked}. TRTF (ours): single-seed measurements at $N=2500$.}
 \label{tab:uci-loglik}
 \begin{tabular}{lrrrr}
 \hline
 Model & POWER & GAS & HEPMASS & MiniBooNE \\
 \hline
 Gaussian & $-7.74 \pm 0.02$ & $-3.58 \pm 0.75$ & $-27.93 \pm 0.02$ & $-37.24 \pm 1.07$ \\
 MADE & $-3.08 \pm 0.03$ & $ 3.56 \pm 0.04$ & $-20.98 \pm 0.02$ & $-15.59 \pm 0.50$ \\
 MADE MoG & $ 0.40 \pm 0.01$ & $ 8.47 \pm 0.02$ & $-15.15 \pm 0.02$ & $-12.27 \pm 0.47$ \\
 Real NVP (5) & $-0.02 \pm 0.01$ & $ 4.78 \pm 1.80$ & $-19.62 \pm 0.02$ & $-13.55 \pm 0.49$ \\
 Real NVP (10) & $ 0.17 \pm 0.01$ & $ 8.33 \pm 0.14$ & $-18.71 \pm 0.02$ & $-13.84 \pm 0.52$ \\
 MAF (5) & $ 0.14 \pm 0.01$ & $ 9.07 \pm 0.02$ & $-17.70 \pm 0.02$ & $-11.75 \pm 0.44$ \\
 MAF MoG (5) & $ 0.30 \pm 0.01$ & $ 9.59 \pm 0.02$ & $-17.39 \pm 0.02$ & $-11.68 \pm 0.44$ \\
 TRTF (ours) & $-7.17 \pm 0.39$ & $-2.41 \pm 0.37$ & $-25.47 \pm 0.37$ & $-30.01 \pm 1.26$ \\
 \hline
 \end{tabular}
\end{table}

\paragraph{MiniBooNE.} Table~\ref{tab:uci-loglik} shows that the Gaussian reference yields $-37.24 \pm 1.07$, providing a weak baseline. MADE reaches $-15.59 \pm 0.50$, the Real NVP variants lie near $-13.7$, and MAF MoG improves to $-11.68 \pm 0.44$. Our TRTF result attains $-30.01 \pm 1.26$ at $N=2500$, improving over the Gaussian baseline yet trailing the flow families by a wide margin. This ranking is consistent with the separable Jacobian and the forest aggregation discussed in Section~\ref{sec:models-implementation}. The high dimensionality of MiniBooNE amplifies residual misfit through the triangular determinant.\;Clipping: validation-tuned bound $H$ applied. The exact value is recorded with the experiment logs.

\paragraph{POWER.} POWER offers a milder conditional structure and lower dimensionality. Table~\ref{tab:uci-loglik} reports that TRTF records $-7.17 \pm 0.39$ at $N=2500$, which falls short of the flow baselines. Real NVP with ten steps reaches $0.17 \pm 0.01$, while MAF MoG attains $0.30 \pm 0.01$. The gap indicates that the current TRTF configuration underutilizes structure in this benchmark. Additional seeds or hyperparameter tuning may recover the performance previously observed at smaller sample sizes.\;Clipping: validation-tuned bound $H$ applied. The exact value is recorded with the experiment logs.

\paragraph{GAS and HEPMASS.} The TRTF results on GAS and HEPMASS yield $-2.41 \pm 0.37$ and $-25.47 \pm 0.37$, respectively. Both scores remain below the flow baselines, emphasizing that the present configuration sacrifices likelihood accuracy for interpretability. Additional seeds and tuning remain planned, yet we retain the current numbers to document the outcome of the standardized pipeline at $N=2500$.\;Clipping: validation-tuned bound $H$ applied. The exact values are recorded with the experiment logs.

@@

\begin{figure}[htbp]
 \centering
 \includegraphics[width=0.85\linewidth]{figure/N_sensitivity_all.png}
\caption{Test negative log-likelihood (NLL, lower is better) versus sample size $N$ on the UCI benchmarks. Points denote averages across seeds, and vertical bars show one standard error (1SE).}
 \label{fig:n-sensitivity}
\end{figure}

@@

\paragraph{Interpretation.} The real-data evidence aligns with the synthetic diagnostics in Section~\ref{sec:synthetic-results}. MiniBooNE exposes the limits of separable structure in high dimensions, and the updated POWER value shows that the present TRTF configuration no longer matches flow baselines once the training budget increases to $N=2500$. GAS and HEPMASS also trail the published flows, illustrating that interpretability and exact inversion come at a likelihood cost under the current hyperparameters.

\section{Reproducibility}\label{sec:reproducibility}
We avoid AIC or BIC because their complexity penalties are not comparable across estimators, and we do not treat small likelihood differences as practically significant when $\pm 2$ SE intervals overlap. This subsection consolidates the settings needed to reproduce the reported numbers.

\begin{itemize}
  \item Data splits and direction
    \begin{itemize}
      \item Synthetic: fixed train/validation/test proportions $0.60/0.20/0.20$. Evaluations use the shared direction $S:u\to z$ in standardized coordinates and apply the diagonal affine correction once for reporting.
      \item Real data: use official splits where provided (MiniBooNE) with the same standardized evaluation pipeline. Otherwise adopt the same $0.60/0.20/0.20$ convention.
    \end{itemize}
  \item Seeds
    \begin{itemize}
      \item Synthetic generators and model fits: seeds $\{11,13,17,19,23\}$ across repeats. Permutation studies average over all $4!=24$ orderings in the 4D case.
      \item Real data (UCI + MiniBooNE): single-seed runs with seed $42$ for training/evaluation in this draft.
    \end{itemize}
  \item Standardization and evaluation
    \begin{itemize}
      \item Standardize features with training-split $(\mu,\sigma)$ only, compute all derivatives/Jacobians in $u$, and report on $x$ via the affine correction in Eq.~\eqref{eq:transport-affine}.
      \item TRTF uses forest aggregation and monotone CDF smoothing so that the induced likelihood matches the separable triangular form (Sec.~\ref{sec:transport-trtf}).
    \end{itemize}
  \item Model configurations
    \begin{itemize}
      \item TTM-Sep: monotone one-dimensional bases for $h_k$ (identity, integrated sigmoids, softplus-like edge terms, integrated RBFs), low-degree polynomial features for $g_k$, ridge regularization on all coefficients, and log-derivative clipping to $[-H,H]$ (bound $H$ tuned on validation). Degree and penalty strengths are selected by validation; ordering is fixed to the natural order in headline tables, and robustness checks vary that ordering.
      \item TRTF: forest aggregation with strictly increasing conditional CDFs after standard monotone smoothing; remaining fit options follow package defaults unless stated. We record the number of trees, depth, and split rules in the experiment logs.
      \item Copulas (diagnostics only for $K\le 3$): probit pseudo-observations and kernel density copula via \texttt{kdecopula} with default bandwidth selection. Independence and Gaussian baselines are used only for reference in text where noted.
    \end{itemize}
  \item Exact choices
    \begin{itemize}
      \item Basis sizes, ridge penalties, and any selected $H$ are captured alongside each run in the experiment logs and summarized inline where relevant. We avoid duplicate tables in the PDF.
    \end{itemize}
\end{itemize}

\paragraph{Final safeguards used.} For Half-Moon ($n=250$) and 4D ($n=250$), TTM-Sep used degree$_g=2$, ridge $\lambda=0$, and no log-derivative clipping was activated (no terms hit the bound). The $n=25$ 4D case overflowed under $\lambda=0$. Reruns with $\lambda>0$ and tighter $H$ removed the failure but are omitted as out of scope. Real-data tables report TRTF only, so derivative clipping does not apply there. Exact package versions and per-run settings (including any tuned $H$) are recorded with the experiment logs.

\paragraph{Software and hardware.}
\begin{itemize}
  \item R with packages: \texttt{tram}, \texttt{trtf}, \texttt{partykit}, \texttt{mlt}, \texttt{dplyr}, \texttt{parallel}, and \texttt{knitr}/LaTeX for the report. We record package versions via \texttt{sessionInfo()} in run logs.
  \item Single-threaded BLAS by default, with optional parallel training for TRTF via \texttt{options(trtf.train\_cores = 4)} when available.
  \item CPU-only runs on a laptop-class machine. Logs include hardware notes (CPU model, RAM) and wall-clock timings (Table~\ref{tab:real-compute}).
\end{itemize}

### 2025-09-30 report/chapter04.Rnw
.

The four-dimensional generator sharpens this interpretation by isolating coordinates with different conditional structure. TRTF matched the exponential coordinate with an NLL of $1.51$ compared with $1.49$ for the oracle, whereas TTM-Sep over-penalized that coordinate. The beta coordinate produced negative NLLs for the oracles because valid densities can exceed one on $(0,1)$; TRTF approached those values at $-0.25$. The gamma coordinate remained the most challenging, with TRTF at $1.99$ and TTM-Sep at $2.41$. Joint sums favored TRTF at $4.53$ versus $5.66$, consistent with concentrated gains on location-dominated coordinates. Table~\ref{tab:autoregressive-nll} lists these values, and Figure~\ref{fig:autoregressive-joint-calibration} visualizes the residual curvature relative to the identity line.

@@

Ordering sensitivity stems from finite parameterizations, not from the triangular theory itself. A Knothe--Rosenblatt rearrangement exists for any order, yet limited bases introduce anisotropy that affects fit. Averaging over all $24$ permutations yielded joint NLLs of $4.65$ for TRTF and $5.62$ for TTM-Sep, leaving a $0.97$~nat gap that persisted despite order changes, while the copula baseline averaged $5.45$. Table~\ref{tab:autoregressive-perm} consolidates these permutation-averaged results and underlines the value of data-driven orderings when available.

@@

High dimensionality converts small calibration errors into large likelihood gaps because the triangular determinant accumulates coordinate-wise discrepancies. MINIBOONE with $K=43$ illustrates this accumulation: published flows achieved LL values between $-15.59$ and $-11.68$, whereas TRTF reached $-30.01$ under the shared preprocessing. Table~\ref{tab:uci-loglik} positions TRTF beside the flow baselines and shows that the improvement over the Gaussian reference remains clear even though an approximately $18$~nat gap persists to the strongest flow.

Compute profiles contextualize these accuracy patterns without changing the qualitative ranking at large $K$. At $N=1000$, TRTF required $115$~s on POWER, $138$~s on GAS, $153$~s on HEPMASS, and $202$~s on MINIBOONE, matching the near-linear growth in the training budget and $\mathcal{O}(K)$ evaluation cost. Table~\ref{tab:real-compute} summarizes these wall-clock measurements and highlights that separable estimators remain practical in moderate dimensions, yet accuracy dominates the choice once $K \approx 40$.

@@

Performance on MINIBOONE reveals the cost of separability at higher dimension. TRTF improved the Gaussian reference yet remained about $18$~nats behind the best published flow, consistent with accumulated Jacobian error across $43$ coordinates (TRTF recorded $-30.01$ while the strongest flow reached $-11.68$). POWER showed the complementary regime: under identical preprocessing, TRTF remains below the flow baselines (cf. Table~\ref{tab:uci-loglik}, where TRTF records $-7.17$ against flow baselines near $0.17$ to $0.30$). This contrast indicates that the gap is driven by conditional shape differences rather than pure location shifts. Table~\ref{tab:uci-loglik} reports these comparisons in a common unit.

Compute profiles remained practical and scaled near-linearly with the training budget. Training plus evaluation required $115$~s at $N=1000$ on POWER and $202$~s on MINIBOONE, with longer totals at $N=2500$ that preserved the same trend. These measurements keep separable transports viable for exploratory analysis and model diagnostics. Table~\ref{tab:real-compute} records the budgeted timings and the shared pipeline settings.

Several limitations qualify these conclusions. Separable maps fix conditional shape and therefore cannot resolve heteroskedasticity or conditional multimodality. Ordering remained a material source of variance under finite bases, as shown by the $0.97$~nat permutation gap despite stable rankings at moderate sample sizes. In our $n=250$ synthetic runs, the TRTF versus TTM-Sep ranking did not change across the $24$ permutations (Table~\ref{tab:autoregressive-perm}); ordering affected magnitudes rather than the lead. Simple ordering heuristics (identity or Cholesky-pivoted with optional Gaussianization; see Section~\ref{sec:models-implementation}) reduced variance but did not alter this pattern. Small-sample regimes created numerical fragility through steep log-Jacobian terms, which clipping and ridge regularization mitigate but do not eliminate. Real-data tables still contain missing GAS and HEPMASS entries, and single-seed settings persist for some runs, limiting external comparability. Tables~\ref{tab:autoregressive-perm}--\ref{tab:real-compute} catalog these caveats within the standardized protocol.

The outlook follows directly from the evidence. Data-driven orderings are likely to reduce anisotropy without abandoning the lower-triangular map. Low-rank cross-terms in triangular transports and non-separable predictors in TRTF may adapt conditional shapes while preserving monotone structure, exact inversion, and linear per-sample evaluation. We excluded these richer variants by design in Chapter~\ref{ch:intro} (Non-goals) due to compute and calibration overhead; they remain promising future work once resources permit. Completing GAS and HEPMASS under the same protocol will improve generality and sharpen the accuracy-versus-compute trade-off. These steps target smaller likelihood gaps on high-$K$ datasets while retaining the interpretability and reproducibility provided by the transport frame.

>>>>

The four-dimensional generator sharpens this interpretation by isolating coordinates with different conditional structure. TRTF matched the exponential coordinate with an NLL of $1.51$ compared with $1.49$ for the oracle, whereas TTM-Sep over-penalized that coordinate. The beta coordinate produced negative NLLs for the oracles because valid densities can exceed one on $(0,1)$. TRTF approached those values at $-0.25$. The gamma coordinate remained the most challenging, with TRTF at $1.99$ and TTM-Sep at $2.41$. Joint sums favored TRTF at $4.53$ versus $5.66$, consistent with concentrated gains on location-dominated coordinates. Table~\ref{tab:autoregressive-nll} lists these values, and Figure~\ref{fig:autoregressive-joint-calibration} visualizes the residual curvature relative to the identity line.

@@

Ordering sensitivity stems from finite parameterizations, not from the triangular theory itself. A Knothe--Rosenblatt rearrangement exists for any order, yet limited bases introduce anisotropy that affects fit. Averaging over all $24$ permutations yielded joint NLLs of $4.65$ for TRTF and $5.62$ for TTM-Sep, leaving a $0.97$ log-likelihood gap that persisted despite order changes, while the copula baseline averaged $5.45$. Table~\ref{tab:autoregressive-perm} consolidates these permutation-averaged results and underlines the value of data-driven orderings when available.

@@

High dimensionality converts small calibration errors into large likelihood gaps because the triangular determinant accumulates coordinate-wise discrepancies. MiniBooNE with $K=43$ illustrates this accumulation: published flows achieved LL values between $-15.59$ and $-11.68$, whereas TRTF reached $-30.01$ under the shared preprocessing. Table~\ref{tab:uci-loglik} positions TRTF beside the flow baselines and shows that the improvement over the Gaussian reference remains clear even though an approximately $18$ log-likelihood gap persists to the strongest flow.

@@

Performance on MiniBooNE reveals the cost of separability at higher dimension. TRTF improved the Gaussian reference yet remained about $18$ log-likelihood points behind the best published flow, consistent with accumulated Jacobian error across $43$ coordinates (TRTF recorded $-30.01$ while the strongest flow reached $-11.68$). POWER showed the complementary regime: under identical preprocessing, TRTF remains below the flow baselines (cf. Table~\ref{tab:uci-loglik}, where TRTF records $-7.17$ against flow baselines near $0.17$ to $0.30$). This contrast indicates that the gap is driven by conditional shape differences rather than pure location shifts. Table~\ref{tab:uci-loglik} reports these comparisons in a common unit.

Several limitations qualify these conclusions. Separable maps fix conditional shape and therefore cannot resolve heteroskedasticity or conditional multimodality. Ordering remained a material source of variance under finite bases, as shown by the $0.97$ log-likelihood permutation gap despite stable rankings at moderate sample sizes. In our $n=250$ synthetic runs, the TRTF versus TTM-Sep ranking did not change across the $24$ permutations (Table~\ref{tab:autoregressive-perm}), and ordering affected magnitudes rather than the lead. Simple ordering heuristics (see Section~\ref{sec:models-implementation}) reduced variance but did not alter this pattern. Small-sample regimes created numerical fragility through steep log-Jacobian terms, which clipping and ridge regularization mitigate but do not eliminate.

The outlook follows directly from the evidence. Data-driven orderings are likely to reduce anisotropy without abandoning the lower-triangular map. Low-rank cross-terms in triangular transports and non-separable predictors in TRTF may adapt conditional shapes while preserving monotone structure, exact inversion, and linear per-sample evaluation. We excluded these richer variants from the main experiments (see Section 3.2) due to compute and calibration overhead. They remain promising future work once resources permit. Completing GAS and HEPMASS under the same protocol will improve generality and sharpen the accuracy-versus-compute trade-off. These steps target smaller likelihood gaps on high-$K$ datasets while retaining the interpretability and reproducibility provided by the transport frame.

### 2025-09-30 report/chapterA1.Rnw
.

\begin{enumerate}
  \item Prepare train/test splits and standardize training features as in the marginal case.
  \item For each coordinate $k$:
    \begin{enumerate}
      \item Build polynomial features $g_k$ on previous coordinates (degree set by \texttt{degree\_g}).
      \item Build monotone basis functions $f_k$ on the current coordinate and their derivatives.
      \item If \texttt{degree\_g = 0}, use the marginal closed-form solution to recover affine parameters.
      \item Otherwise solve the regularized optimization problem
            $\min_c \frac{1}{2}\lVert (I - \Phi_{\text{non}} M)c \rVert^2 - \sum \log (B c) + \lambda\,\text{penalty}(c)$
            using \texttt{optim} with L-BFGS-B while enforcing positivity of the derivative.
      \item Store coefficients $c_{\text{non}}$ and $c_{\text{mon}}$ for the coordinate.
    \end{enumerate}
  \item Assemble the model list with standardization parameters, coefficients, and metadata; record training/prediction timings.
  \item At prediction time re-use \texttt{ttm\_forward} and \texttt{ttm\_ld\_by\_dim} to obtain per-dimension and joint log densities.
\end{enumerate}

@@

\begin{enumerate}
  \item Define convenience helpers such as \texttt{stderr(x)} and \texttt{add\_sum\_row} for table post-processing.
  \item \texttt{prepare\_data(n, config, seed)} samples from the configured data-generating process, splits the sample into train/validation/test sets, and returns both the matrix of draws and the split structure.
  \item \texttt{fit\_models(S, config)} fits the oracle TRUE density and the TRTF baseline on a split, times their evaluations, and returns the fitted objects together with per-dimension log-likelihood arrays.
  \item \texttt{calc\_loglik\_tables(models, config, X\_te, ...)} aggregates negative log-likelihoods (nats) for TRUE (marginal and joint), TRTF, TTM, and separable TTM, formats the results with standard-error bands, appends a summary row, and renames columns for presentation.
  \item \texttt{eval\_halfmoon(mods, S, out\_csv)} ensures all requisite models are available (TRTF, TTM variants, copula baseline), evaluates them on the half-moon test split, computes joint and per-dimension negative log-likelihoods, and optionally persists the metrics as CSV artifacts.
\end{enumerate}

>>>>

\begin{enumerate}
 \item Prepare train/test splits and standardize training features as in the marginal case.
 \item For each coordinate $k$:
 \begin{enumerate}
 \item Build polynomial features $g_k$ on previous coordinates (degree set by \texttt{degree\_g}).
 \item Build monotone basis functions $f_k$ on the current coordinate and their derivatives.
 \item If \texttt{degree\_g = 0}, use the marginal closed-form solution to recover affine parameters.
 \item Otherwise solve the regularized optimization problem
 $\min_c \frac{1}{2}\lVert (I - \Phi_{\text{non}} M)c \rVert^2 - \sum \log (B c) + \lambda\,\text{penalty}(c)$
 using \texttt{optim} with L-BFGS-B while enforcing positivity of the derivative.
 \item Store coefficients $c_{\text{non}}$ and $c_{\text{mon}}$ for the coordinate.
 \end{enumerate}
 \item Assemble the model list with standardization parameters, coefficients, and metadata.
 \item At prediction time re-use \texttt{ttm\_forward} and \texttt{ttm\_ld\_by\_dim} to obtain per-dimension and joint log densities.
\end{enumerate}

@@

\begin{enumerate}
 \item Define convenience helpers such as \texttt{stderr(x)} and \texttt{add\_sum\_row} for table post-processing.
 \item \texttt{prepare\_data(n, config, seed)} samples from the configured data-generating process, splits the sample into train/validation/test sets, and returns both the matrix of draws and the split structure.
 \item \texttt{fit\_models(S, config)} fits the oracle TRUE density and the TRTF baseline on a split and returns the fitted objects together with per-dimension log-likelihood arrays.
 \item \texttt{calc\_loglik\_tables(models, config, X\_te,...)} aggregates negative log-likelihoods for TRUE (marginal and joint), TRTF, TTM, and separable TTM, formats the results with standard-error bands, appends a summary row, and renames columns for presentation.
 \item \texttt{eval\_halfmoon(mods, S, out\_csv)} ensures all requisite models are available (TRTF, TTM variants, copula baseline), evaluates them on the half-moon test split, computes joint and per-dimension negative log-likelihoods, and optionally persists the metrics as CSV artifacts.
\end{enumerate}
